\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-06-07 doc/paper-jss/sdmTMB-paper.Rnw
Note R and Matrix versions in timing
Edits to timing section

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
Our test was restricted to one core and default \R\ algebra libraries.
% \sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
% These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
\added{
With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three- to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
At larger sample sizes, \pkg{inlabru} was more affected by mesh resolution than \sdmTMB\ (Figure~\ref{fig:timing}c).
\pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\sim$ 200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but can only fit spatial GMRFs.
At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
}
Speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.

>>>>

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
Our test was restricted to one core and default \R\ algebra libraries \added{using \R\ 4.4.0 and \pkg{Matrix} version 1.7.0.
With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three- to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
At larger sample sizes, \pkg{inlabru} was more affected by mesh resolution than \sdmTMB\ (Figure~\ref{fig:timing}c).
\pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\approx$~200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
The \pkg{mgcv} model was fit with the \fct{bam} function instead of the standard \fct{gam}.
The \fct{bam} function here uses numerical methods optimized for large datasets, including a discretization of covariate values \citep{wood2015, wood2017b, li2020a}.
The standard \fct{gam} function was approximately 60 times slower than \fct{bam} for the largest dataset with the most complex mesh (not shown).
\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but to our knowledge can only fit spatial GMRFs.
At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
}
Speed increases can allow for more rapid and thorough model exploration and experimentation with this class of computationally intensive models.
For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.
